---
title: "Laboratory Animals Code"
author: "Angela Jeffers"
date: "12/21/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Steps 1-2: Establish study objectives and collect pilot data. These steps are performed prior to the simulation code.

## Step 3: Verify assumptions and specify simulation parameters.

## Summary statistics for each endpoint based on pilot data from four National Toxicology Program (NTP) 90-day toxicology studies are documented here. These summary statistics are used to define the simulation parameters. The assumption of normality for these endpoints was previously verified during the pilot data phase using Q-Q plots and histograms.

```{r Load packages}
## Load libraries necessary for running the simulation and power analysis.
library(tidyverse)
library(doParallel) # For parallel processing - can execute foreach loops created in the Parallel package.
library(foreach)    # Provides looping construct for executing R code repeatedly.
library(clinfun)    # Jonckheere trend test is in this package.
library(scales)     # The percent function is in this package.
```

```{r Defining simulation parameters}
# After pilot data is sufficiently reviewed, specify the simulation study parameters.

# Create a tibble that contains the endpoints of interest and their corresponding values. In this case, we are interested in male liver and testis weights. This tibble will be referenced in the simulation and power analysis code. 

sim<-tibble(
  endpoint=c("Testis Weight", "Male Liver Weight"),
  mu=c(2.08, 17.3),# Mean of control groups per endpoint (testis weights, liver weights)
  sd=c(.13, 1.42), # Define the standard deviation per endpoint (testis weights, liver weights)
  ndose=10 # Number of animals per dose group - the same for control and treated data
  )

N=10000 # Number of tests to run; adjust as needed. This equates to Nsim. 
nControl=10 # Number per dose group ('AnimalID' = 1-10)
nTotal=40 # Total number of animals
iter=1:N # Total number of iterations
TrtDoseGrps=c(1, 2, 3) # Dose group values for treated data. There are 4 total dose groups including the control.
Simulation=c("15% Effect") # Label for the effect size
```

## Here, we set up data frames for the simulation study and simulate data for both endpoints. The control and treated data are combined into a single data frame. 

```{r Set up control data}
# Note on the code: rep(value, # repeats); The rep() function repeats characters.
# Note on the code: expand.grid() function  creates a data frame of all of the combinations of vectors passed as argument (in this case, for each endpoint, AnimalID and effect size).
# We need to run the simulation for each animal per iteration (i.e., iteration 1 will have 10 observations for each endpoint and effect size combination).

# We will simulate the control and treated data separately.
# First, we set up the blank control data results.
# EffectSize is set to 1 since the control group is never affected by the effect size.
 controlresults= as_tibble(expand.grid(Endpoint=sim$endpoint, Iteration=iter, dose=0, AnimalID=1:nControl, Simulation=Simulation, EffectSize=1)) 

# Sort by endpoint and iteration for easier data processing and viewing
controlresults<- controlresults %>%
  arrange(Endpoint, Simulation, Iteration)
```

```{r Set up treated data}
# Once control data is simulated, Simulate the treated data. Start by setting up a blank results tibble.
 treatedresults= as_tibble(expand.grid(Endpoint=sim$endpoint, Iteration=iter, dose=TrtDoseGrps, AnimalID=1:nControl, Simulation=Simulation, EffectSize=NA, sample=NA))
 # Sort by endpoint and iteration for data processing/viewing
 treatedresults<- treatedresults %>%
   arrange(Endpoint, Simulation, Iteration)
 
# Create effect size column used to multiply means by for simulating data. 
# Linear interpolation was used to determine the effect sizes for each dose group.
# Simulation is based on the effect size in the highest dose group
 treatedresults = treatedresults %>% mutate(EffectSize = case_when(
   Simulation == "15% Effect" & dose == 1 ~ .95,
   Simulation == "15% Effect" & dose == 2 ~ .9,
   Simulation == "15% Effect" & dose == 3 ~ .85,
   TRUE ~ 0)
 )
 
# Create STE Multiplier column - 2-fold increase in the SD for the highest dose group; proportional increase across dose groups.
# Increase by .333 per dose group in treated data. This indicates heteroscedasticity. 
 treatedresults = treatedresults %>% mutate(STESim = case_when(
   Simulation == "No Effect" ~ 1, 
   dose == 0 ~ 1,
   dose == 1 ~ 1.333,
   dose == 2 ~ 1.667,
   dose == 3 ~ 2,
   TRUE ~ 0)
 )
```

## Step 4: Simulate data.

```{r Simulate control and treated data}
set.seed(1563) # Set the seed in the code for reproducibility.

# Simulate control data
   controlresults$sample=rnorm(nrow(controlresults),
                              mean=rep(c(sim$mu), each=nrow(controlresults)/2),
                              sd=rep(c(sim$sd), each=nrow(controlresults)/2))
# SD multiplier column for control data
 controlresults = controlresults %>% mutate(STESim = case_when(
      Simulation == "No Effect" ~ 1,
      dose == 0 ~ 1
 ))
 

# Simulate treated data
treatedresults$sample=rnorm(nrow(treatedresults),
                             mean=treatedresults$EffectSize*rep(c(sim$mu), each=(nrow(treatedresults)/2)),
                             sd=treatedresults$STESim*rep(c(sim$sd), each=(nrow(treatedresults)/2)))
  
# Note: The 'each' part of the code is necessary to tell the code how many simulations to run for each endpoint. There are 2 endpoints, so the simulation is repeated 2 times.
# Note: Repeat for each mean and effect size 
```

```{r Combine control and treated data}
# Merge control and treated data - set up for statistical analysis results with empty column values
 combinedresults<-rbind(controlresults, treatedresults) %>%
  mutate(ndose=10)
## Sort by endpoint 
  combinedresults<-combinedresults %>% 
    arrange(Endpoint, Simulation, Iteration, dose)
```

## Step 5: Perform statistical testing.
## Step 6: Repeat Nsim times.
## Step 7: Calculate Power.

```{r}
# Takes approximately 10 minutes to run this portion of the code.

endpoint_list = c("Testis Weight", "Male Liver Weight")
power_results = c()

# Repeat for each endpoint.
for (edpt in 1:length(endpoint_list)){
  T = 0
  # Repeat Nsim times (N=10,000).
  for (i in 1:N){
    oneRepDataset <- combinedresults %>%
      filter(Endpoint == endpoint_list[edpt],
             Iteration == i)
    # Perform statistical testing.
    # Create a tally of the number of H0 rejections.
    Signif_Test = jonckheere.test(oneRepDataset$sample, oneRepDataset$dose)$p.value < 0.05
    if(Signif_Test){
      T = T + 1
    }
    if(i%%1000 == 1){
      print(i)
    }
  }
  # Calculate power (after all Nsim calculations for each endpoint are completed)
  power_results = as.data.frame(rbind(power_results, c(endpoint_list[edpt], percent((T/N), accuracy=.01))))
  names(power_results) = c("Endpoint", "Power")
}

power_results
```

## Step 8: Repeat.